---
title: "Script for the study of the freshwater ecosystems at Toros-Menalla"
author: "Axelle Gardin"
date: 17/12/2025
format: 
  html:
    toc: true
    code-fold: show
    number-sections: true
    fig-cap-location: bottom
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This Quarto document provides the R code used for the analyses and figures presented in the manuscript:

**"Reconstructing the freshwater paleoecosystem diversity of Toros-Menalla (Late Miocene, Chad) from an integrated faunal perspective"** by Axelle Gardin, Olga Otero, Géraldine Garcia, Clarisse Nekoulnang, Fabrice Lihoreau, Abderamane Moussa, Mathieu Schuster, Lorenzo Scribano, & Franck Guy.

This work focuses on reconstructing Late Miocene freshwater habitat at Toros-Menalla (northern Chad Basin), where aquatic and terrestrial ecosystems were tightly interconnected. While terrestrial vertebrate assemblages have been extensively studied, aquatic habitats have remained poorly resolved.

By integrating data from five Toros-Menalla sites (including fossil freshwater-dwelling vertebrate assemblages, and discussion with depositional contexts, stable isotopes, and field observations) this study highlights a complex network of freshwater habitats, ranging from floodplains and swamps to open waterbodies with currents. Most assemblages represent spatial habitat diversity rather than temporal succession, with the exception of TM266, which may reflect vertical mixing. This work emphasizes the importance of systematic, grid-based fossil collection and sieving to capture representative aquatic biodiversity and reconstruct paleoecosystems with ecological precision.

## Objectives of this script

The code includes:

-   Calculation of standardized taxonomic ranks of abundance;
-   Generation of rarefaction curves based on the number of specimens;
-   Identification of sampling thresholds from rarefaction curves;
-   Visualisation of the sampling effort among clades with rarefaction curves;
-   MNI sensitivity analysis;
-   Hierachical clustering of asemblages;
-   PCA ordination;
-   Calculation of ecological scores;
-   Visualisation of the assemblage structure across sites and comparison of proportions of aquatic vs terrestrial vertebrate taxa across the five studied sites from Toros-Menalla.

## About the data

The datasets are:

-   `TM_Aquatic_NISP_MNI.csv` – contains abundance data for freshwater-dwelling taxa in the five studied sites (TM90, TM242, TM252, TM266, TM337), including **Number of Identifierd Specimens (NIPS)** and **Minimum Number of Individuals** (**MNI)**;
-   `TM_aquatic_terrestrial_proportion.csv` – contains proportions of aquatic and terrestrial vertebrate groups used to build comparative pie charts.
-   `TM_taxon_ecological_matrix.csv` – contains ecological categories (habitat association and ecological tolerance) associated to each taxon, according to Table 1 of the main text.

## Citation

***This repository is licensed under the CC BY 4.0 License.*** If you use the data, code, or figures from this repository in your work, please cite it as:

Axelle Gardin, Olga Otero, Géraldine Garcia, Clarisse Nekoulnang, Fabrice Lihoreau, Abderamane Moussa, Mathieu Schuster, Lorenzo Scribano, & Franck Guy (2026). *Toros-Menalla Freshwater Ecosystems*. GitHub repository. <https://doi.org/10.5281/zenodo.17963310>

------------------------------------------------------------------------

## Set up workspace

### Load required libraries

```{r}
library(dplyr)
library(ggplot2)
library(ggplotify)
library(ggrepel)
library(gridExtra)
library(iNEXT)
library(patchwork)
library(pheatmap)
library(plotly)
library(purrr)
library(pvclust)
library(tibble)
library(tidyr)
library(vegan)
```

------------------------------------------------------------------------

### Load required datasets

```{r}
# TM_Aquatic_NISP_MNI
TM_data <- read.table(file="data/TM_Aquatic_NISP_MNI.csv", head=TRUE, dec=".", sep=";", row.names=1)

axa <- row.names(TM_data)
sites <- c("TM90", "TM242", "TM252", "TM266", "TM337")

# NISP
NISP <- TM_data[, 1:5]
colnames(NISP) <- sites

# MNI
MNI <- TM_data[, 6:10]
colnames(MNI) <- sites

# Aquatic vs terrestrial proportion 
TM_AqTer <- read.table(file="data/TM_aquatic_terrestrial_proportion.csv", head=TRUE, dec=".", sep=";", row.names=1)

# Ecological matrix
eco_matrix <- read.table(file="data/TM_taxon_ecological_matrix.csv", head=TRUE, dec=".", sep=";", row.names=1)

```

### Calculation of standardized taxonomic ranks of abundance

The following code calculates a standardized taxonomic rank of abundance for each taxon within each site, following the method described by Le Fur et al. (2014).

**Reference:** Le Fur, Soizic, et al. "Toros-Menalla (Chad, 7 Ma), the earliest hominin-bearing area: How many mammal paleocommunities?" *Journal of Human Evolution* 69 (2014): 79-90.

```{r}
compute_standardized_ranks <- function(x) {
    # Create a result vector initialized with 0 (for absent taxa)
  r <- rep(0, length(x))
    # Identify indices of present taxa (abundance > 0)
  nonzero_idx <- which(x > 0)
    # If all taxa are absent, return the vector of zeros
  if(length(nonzero_idx) == 0) return(r)
    # Extract non-zero values
  values <- x[nonzero_idx]
    # Sort unique values in decreasing order (highest abundance first)
  unique_vals <- sort(unique(values), decreasing = TRUE)
    # Assign an abundance rank to each taxon based on its value
  unique_ranks <- match(values, unique_vals)
    # Maximum rank value
  max_unique <- length(unique_vals)
    # Standardize the inverse rank (most abundant = 1.0, least = 1/max_unique)
  std_ranks <- (max_unique - unique_ranks + 1) / max_unique
    # Assign standardized ranks back to the corresponding taxa
  r[nonzero_idx] <- std_ranks
  return(r)
}

# Apply the function to each site (each column of MNI)
std_ranks_df <- as.data.frame(apply(MNI, 2, compute_standardized_ranks))
rownames(std_ranks_df) <- taxa

write.csv(std_ranks_df, "results/TM_std_ranks_abundance.csv", row.names = FALSE)
```

# Generation of rarefaction curves based on the number of specimens

```{r}
# Rarefaction and extrapolation based on abundance data
DataInfo(NISP, datatype = "abundance")
cor(DataInfo(NISP)$S.obs, DataInfo(NISP)$n)

# Generate rarefaction curves
D_abund <- iNEXT(NISP, datatype = "abundance")

png("figures/Fig2_rarefaction_curves.png", width = 2000, height = 1500, res = 300)
plot(D_abund, xlim = c(0, 2300))
dev.off()
```

## Identification of sampling thresholds from rarefaction curves

To assess sampling sufficiency, we compute rarefaction and extrapolation curves using the iNEXT package. We define a diversity plateau as the point where the marginal gain in taxonomic diversity falls below 0.01 species per additional individual. Beyond this point, diversity increases very slowly: more than 100 additional individuals would be required, on average, to detect one new taxon.

```{r}

# Extract estimated diversity values
df1 <- D_abund$iNextEst[[1]]

# Calculate the slope between successive points
slope_df <- df1 %>%
  filter(Method != "Observed") %>%
  arrange(m) %>%
  group_by(Assemblage) %>%
  mutate(
    delta_m = lead(m) - m,
    delta_qD = lead(qD) - qD,
    slope = delta_qD / delta_m,
    plateau = ifelse(abs(slope) < 0.01, TRUE, FALSE)
  )

# Identify the first point where the slope drops below 0.01
plateau_points <- slope_df %>%
  filter(plateau) %>%
  slice(1) %>%
  select(Assemblage, m, qD, slope)

# Display the plateau values per assemblage
plateau_points

```

## Visualisation of the sampling effort among clades with rarefaction curves

*This script generates Figure S1 of the Supplementary Information.*

To evaluate sampling effort across the main clades, we used specimen-based rarefaction curves computed separately for each clade and for each site. These curves provide a first-order assessment of whether taxonomic richness approaches an asymptote within clades, and allow comparison of relative sampling completeness among groups and sites. However, this approach has intrinsic limitations. In particular, part of the crocodylian material consists of isolated, morphologically non-diagnostic teeth that likely belong to *Euthecodon* or gavialoid taxa, but cannot be confidently assigned beyond Crocodylia indet. These remains are therefore excluded from the present dataset. As a consequence, rarefaction curves suggest an apparently good sampling of Crocodylia, while masking a higher underlying crocodylian diversity that is known to exist but cannot be resolved taxonomically. Rarefaction curves are thus informative about sampling effort within the identified assemblage, but cannot capture hidden diversity linked to taxonomic uncertainty.

```{r}
df <- NISP %>%
  tibble::rownames_to_column("taxon") %>%
  pivot_longer(
    cols = -taxon,
    names_to = "site",
    values_to = "n"
  )

map <- tribble(
  ~taxon,       ~clade,
  "Alestinae",  "Characiformes",
  "Bagridae",   "Siluriformes",
  "Cichlidae",  "Acanthomorpha",
  "Clariidae",  "Siluriformes",
  "Claroteidae","Siluriformes",
  "Cyprinidae", "Cypriniformes",
  "Gymnarchus", "Osteoglossiformes",
  "Crocodylus", "Crocodylia",
  "Euthecodon", "Crocodylia",
  "Gavialoidea","Crocodylia",
  "Tetraodontidae", "Tetraodontiformes",
  "Trionychidae", "Testudines",
  "Heterotis", "Osteoglossiformes",
  "Hexaprotodon", "Semi-aquatic Mammals",
  "Hydrocynus", "Characiformes",
  "Lates", "Acanthomorpha",
  "Libycosaurus", "Semi-aquatic Mammals",
  "Mochokidae", "Siluriformes",
  "Mormyridae", "Osteoglossiformes",
  "Polypterus", "Polypteriformes",
  "Semlikiichthys", "Acanthomorpha",
  "Synodontis", "Siluriformes"
)

df <- df %>%
  left_join(map, by = "taxon")

clade_colors <- c(
  "Characiformes" = "#B5D2EA",
  "Siluriformes" = "#2B3E5A",
  "Acanthomorpha" = "#8DACC6",
  "Cypriniformes" = "#4D5860",
  "Osteoglossiformes" = "#577591",
  "Polypteriformes" = "#0B1319",
  "Tetraodontiformes" = "#12679B",
  "Testudines" = "#897E51",
  "Crocodylia" = "#ADA379",
  "Semi-aquatic Mammals" = "#A6786F"
)
df$site <- factor(df$site)

prep_inext_site <- function(data_site) {
  data_site %>%
    group_by(clade) %>%
    summarise(abundances = list(setNames(n, taxon))) %>%
    { setNames(.$abundances, .$clade) }
}

results <- list()
sites <- as.character(unique(df$site))

for(s in sites) {
  cat("Processing site:", s, "\n")
  data_site <- df %>% filter(site == s & n > 0)
  data_list <- prep_inext_site(data_site)
  out <- iNEXT(data_list, datatype = "abundance")
  results[[s]] <- out
}

#  Build and save graph for each site
## TM90
    ggiNEXT(results$TM90, type = 1) +
    ggtitle(paste("Rarefaction per clade - TM90")) +
    xlim(0, 100) +
    xlab("Number of specimens") +
    ylab("Taxon richness") +
    scale_colour_manual(values = clade_colors) +
    scale_fill_manual(values = clade_colors) +
      theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

## TM242
    ggiNEXT(results$TM242, type = 1) +
    ggtitle(paste("Rarefaction per clade - TM242")) +
    xlim(0, 100) +
    xlab("Number of specimens") +
    ylab("Taxon richness") +
    scale_colour_manual(values = clade_colors) +
    scale_fill_manual(values = clade_colors) +
      theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

## TM252
    ggiNEXT(results$TM252, type = 1) +
    ggtitle(paste("Rarefaction per clade - TM252")) +
    xlim(0, 100) +
    xlab("Number of specimens") +
    ylab("Taxon richness") +
    scale_colour_manual(values = clade_colors) +
    scale_fill_manual(values = clade_colors) +
      theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

## TM266
    ggiNEXT(results$TM266, type = 1) +
    ggtitle(paste("Rarefaction per clade - TM266")) +
    xlim(0, 100) +
    xlab("Number of specimens") +
    ylab("Taxon richness") +
    scale_colour_manual(values = clade_colors) +
    scale_fill_manual(values = clade_colors) +
      theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

## TM337
    ggiNEXT(results$TM337, type = 1) +
    ggtitle(paste("Rarefaction per clade - TM337")) +
    xlim(0, 100) +
    xlab("Number of specimens") +
    ylab("Taxon richness") +
    scale_colour_manual(values = clade_colors) +
    scale_fill_manual(values = clade_colors) +
      theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
    
# Building figure S1 with all sites combined
p_TM90 <- ggiNEXT(results$TM90, type = 1) +
  ggtitle("TM90") +
  xlim(0, 100) +
  xlab("Number of specimens") +
  ylab("Taxon richness") +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

p_TM242 <- ggiNEXT(results$TM242, type = 1) +
  ggtitle("TM242") + 
  xlim(0, 100) +
  xlab("Number of specimens") +
  ylab("Taxon richness") +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

p_TM252 <- ggiNEXT(results$TM252, type = 1) +
  ggtitle("TM252") + 
  xlim(0, 100) +
  xlab("Number of specimens") +
  ylab("Taxon richness") +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

p_TM266 <- ggiNEXT(results$TM266, type = 1) +
  ggtitle("TM266") + 
  xlim(0, 100) +
  xlab("Number of specimens") +
  ylab("Taxon richness") +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

p_TM337 <- ggiNEXT(results$TM337, type = 1) +
  ggtitle("TM337") + 
  xlim(0, 100) +
  xlab("Number of specimens") +
  ylab("Taxon richness") +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

# Extract the legend
legend_plot <- ggiNEXT(results$TM90, type = 1) +
  scale_colour_manual(values = clade_colors) +
  scale_fill_manual(values = clade_colors) +
  theme_minimal() +
  theme(legend.position = "right")

get_legend <- function(p) {
  g <- ggplotGrob(p)
  leg <- g$grobs[which(sapply(g$grobs, function(x) x$name) == "guide-box")]
  return(leg[[1]])
}

legend_only <- get_legend(legend_plot)
legend_plot_for_patchwork <- as.ggplot(legend_only)
 
# Combine graphs
combined <- (p_TM90 | p_TM242) /
            (p_TM252 | p_TM266) /
            (p_TM337 | legend_plot_for_patchwork)

ggsave("figures/FigS1_rarefaction_per_clades.png",
       plot = combined, width = 8, height = 13)

```

# MNI sensitivity analysis

*This script generates Tables S1 to S4 of the Supplementary Information.*

MNI was estimated based on the most abundant skeletal element, which varies among sites, taxa, and, when observed, sex and size classes. For fish taxa primarily represented by isolated teeth, MNI estimates were derived from tooth counts in modern relative species: 15-25 teeth for Alestidae (Vienne et al. 2026), 15-30 for *Hydrocynus* (Vienne et al. 2026), 22 for Labeo (Pasco-Viel et al. 2014), 34-44 for *Gymnarchus* (Bigorne, 2003), and four beak elements for *Tetraodon* (Lévêque and Paugy, 2003). Where modern data indicate substantial variation in the number of teeth per individual, we used a range rather than a single value. For assemblages in which the plausible range of teeth per individual resulted in different MNI estimates (all except TM242), we conducted a sensitivity analysis of the standardised ranks of abundance.

Overall, it indicates that the relative abundance patterns and associated ecological interpretations are robust to the plausible uncertainty in MNI estimates (see supplementary material). For the remainder of the study, we use the lowest MNI estimate to remain conservative. We did not observe worn or digested teeth, suggesting that the contribution of shed teeth to our estimates is negligible.

**Generate MNI scenarios combining minimum and maximum estimates for selected taxa**

```{r}
make_scenario <- function(MNI, site, Alestinae = NULL, Hydrocynus = NULL, Gymnarchus = NULL) {
  MNI_new <- MNI
    if (!is.null(Alestinae)) {MNI_new["Alestinae", site] <- Alestinae}
    if (!is.null(Hydrocynus)) {MNI_new["Hydrocynus", site] <- Hydrocynus}
    if (!is.null(Gymnarchus)) {MNI_new["Gymnarchus", site] <- Gymnarchus}
    return(MNI_new) }

#TM90: MNI for Alestinae is estimated to vary from 7 to 9
TM90_scenarios <- list(min = make_scenario(MNI, site = "TM90", Alestinae = 7), max = make_scenario(MNI, site = "TM90", Alestinae = 9))

# TM242: MNI remains unchanged

# TM252: MNI for Alestinae is estimated to vary from 74 to 125, and from 11 and 15 for Hydrocynus.
TM252_scenarios <- list(min_min = make_scenario(MNI, site = "TM252", Alestinae = 74, Hydrocynus = 11), min_max = make_scenario(MNI, site = "TM252", Alestinae = 74, Hydrocynus = 15),max_min = make_scenario(MNI, site = "TM252", Alestinae = 125, Hydrocynus = 11), max_max = make_scenario(  MNI, site = "TM252", Alestinae = 125, Hydrocynus = 15))

#TM266: MNI for Hydrocynus is estimated to vary from 11 to 25, and from 5 and 6 for Gymnarchus
TM266_scenarios <- list(min_min = make_scenario(MNI, site = "TM266", Hydrocynus = 11, Gymnarchus = 5), min_max = make_scenario(MNI, site = "TM266", Hydrocynus = 11, Gymnarchus = 6), max_min = make_scenario(MNI,site = "TM266", Hydrocynus = 25, Gymnarchus = 5), max_max = make_scenario(MNI, site = "TM266", Hydrocynus = 25, Gymnarchus = 6))

#TM337: MNI for Alestinae is estimated to vary from 11 to 16
TM337_scenarios <- list(min = make_scenario(MNI, site = "TM337", Alestinae = 11 ), max = make_scenario(MNI, site = "TM337", Alestinae = 16))
```

**Recalculate raw abundance ranks for each MNI scenario**

```{r}
#Calculate ranks of abundance (not standardised)
compute_ranks <- function(x) {
  r <- rep(0, length(x))
  nonzero_idx <- which(x > 0)
  if(length(nonzero_idx) == 0) return(r)
  values <- x[nonzero_idx]
  unique_vals <- sort(unique(values), decreasing = TRUE)
  unique_ranks <- match(values, unique_vals)
  r[nonzero_idx] <- unique_ranks
  return(r)
}

# Calculate ranks from the original MNI dataframe
ranks_df <- as.data.frame(apply(MNI, 2, compute_ranks))

rownames(ranks_df) <- taxa
colnames(ranks_df) <- sites

# Calculate alternative ranks in the scenarios for each site
calculate_ranks <- function(MNI_matrix) {
  ranks <- as.data.frame(apply(MNI_matrix, 2, compute_ranks))
  rownames(ranks) <- rownames(MNI_matrix)
  colnames(ranks) <- colnames(MNI_matrix)
  return(ranks)
}
```

**Compare abundance ranks across MNI scenarios**

```{r}
#TM90: some ranks vary between 2 and 1, but does not affect the main taxa
TM90_ranks <- lapply(TM90_scenarios, calculate_ranks)
TM90_comparison <- lapply(TM90_ranks, function(x) {x[, "TM90"] - ranks_df[, "TM90"]})
TM90_comparison

TM90_comp_ranks <- data.frame(Taxon = taxa, MNI_reference = MNI[, "TM90"], 
MNI_min = TM90_scenarios$min[, "TM90"], MNI_max = TM90_scenarios$max[, "TM90"], Rank_reference = ranks_df[, "TM90"], Rank_min = TM90_ranks$min[, "TM90"], Rank_max = TM90_ranks$max[, "TM90"])
TM90_comp_ranks

#TM252: Ranks are not affected by alternative MNI.
TM252_ranks <- lapply(TM252_scenarios, calculate_ranks) 
TM252_comparison <- lapply(TM252_ranks, function(x) {x[, "TM252"] - ranks_df[, "TM252"]})
TM252_comparison

TM252_comp_ranks <- data.frame(Taxon = taxa, MNI_reference = MNI[, "TM252"], MNI_min_min = TM252_scenarios$min_min[, "TM252"],MNI_min_max = TM252_scenarios$min_max[, "TM252"], MNI_max_min = TM252_scenarios$max_min[, "TM252"], MNI_max_max = TM252_scenarios$max_max[, "TM252"], Rank_reference = ranks_df[, "TM252"], Rank_min_min = TM252_ranks$min_min[, "TM252"], Rank_min_max = TM252_ranks$min_max[, "TM252"], Rank_max_min = TM252_ranks$max_min[, "TM252"], Rank_max_max = TM252_ranks$max_max[, "TM252"])
TM252_comp_ranks

#TM266: some ranks vary between 2 and 1, but does not affect the main taxa
TM266_ranks <- lapply(TM266_scenarios, calculate_ranks)
TM266_comparison <- lapply(TM266_ranks, function(x) {x[, "TM266"] - ranks_df[, "TM266"]})
TM266_comparison

TM266_comp_ranks<-data.frame(Taxon = taxa, MNI_reference = MNI[, "TM266"], MNI_min_min = TM266_scenarios$min_min[, "TM266"], MNI_min_max = TM266_scenarios$min_max[, "TM266"], MNI_max_min = TM266_scenarios$max_min[, "TM266"], MNI_max_max = TM266_scenarios$max_max[, "TM266"], Rank_reference = ranks_df[, "TM266"], Rank_min_min = TM266_ranks$min_min[, "TM266"], Rank_min_max = TM266_ranks$min_max[, "TM266"], Rank_max_min = TM266_ranks$max_min[, "TM266"], Rank_max_max = TM266_ranks$max_max[, "TM266"])

#TM337: some ranks vary between 2 and 1, but does not affect the main taxa
TM337_ranks <- lapply(TM337_scenarios, calculate_ranks)
TM337_comparison <- lapply(TM337_ranks,function(x) {x[, "TM337"] - ranks_df[, "TM337"]})
TM337_comparison

TM337_comp_ranks <- data.frame(Taxon = taxa, MNI_reference = MNI[, "TM337"], MNI_min = TM337_scenarios$min[, "TM337"], MNI_max = TM337_scenarios$max[, "TM337"], Rank_reference = ranks_df[, "TM337"], Rank_min = TM337_ranks$min[, "TM337"], Rank_max = TM337_ranks$max[, "TM337"])
TM337_comp_ranks

write.csv(TM90_comp_ranks, "results/TableS1_TM90_MNI_sensitivity_test.csv", row.names = FALSE)
write.csv(TM252_comp_ranks, "results/TableS2_TM252_MNI_sensitivity_test.csv", row.names = FALSE)
write.csv(TM266_comp_ranks, "results/TableS3_TM266_MNI_sensitivity_test.csv", row.names = FALSE)
write.csv(TM337_comp_ranks, "results/TableS4_TM337_MNI_sensitivity_test.csv", row.names = FALSE)
```

# Hierarchical clustering of assemblages

Similarity in taxonomic structure among assemblages was explored using hierarchical cluster analysis based on the Bray-Curtis dissimilarity index. The Bray-Curtis index quantifies dissimilarity in standardised taxonomic rank abundance, ranging from 0 (identical assemblages) to 1 (no taxa in common). Ranks were Hellinger-transformed before clustering to reduce the influence of highly abundant taxa. Given the small number of assemblages (n = 5), the resulting clusters are considered exploratory, as their robustness cannot be reliably assessed statistically.

```{r}

std_ranks_df_long <- t(std_ranks_df)

bray_curtis <- vegdist(std_ranks_df_long, method = "bray")
hc <- hclust(bray_curtis, method = "average")

# Visualize dendrogram
png(filename = "figures/Fig3_hierarchical_clustering.png",
    width = 600, height = 400)

plot(hc, main = "Hierarchical clustering (Bray–Curtis)", xlab = "", sub = "")

dev.off()
```

# PCA ordination

A principal component analysis was performed on the matrix of standardized ranks of abundance to explore patterns of taxonomic structure among assemblages. The contributions of taxa to each principal component were calculated from their squared loadings.

```{r}

# PCA on standardized abundance ranks
pca_ranks <- prcomp(
  std_ranks_df_long,
  center = TRUE,
  scale. = FALSE
)

summary(pca_ranks)

# Select top 10 taxa contributing to PC
sites <- as.data.frame(pca_ranks$x[, 1:3])
sites$Site <- rownames(sites)

loadings <- as.data.frame(pca_ranks$rotation[, 1:3])
loadings$Taxon <- rownames(loadings)

# PC1-PC2
loadings$Magnitude_PC12 <- sqrt(loadings$PC1^2 + loadings$PC2^2)

top10_PC12 <- loadings[order(loadings$Magnitude_PC12, decreasing = TRUE), ][1:10, ]

# PC1-PC3
loadings$Magnitude_PC13 <- sqrt(loadings$PC1^2 + loadings$PC3^2)

top10_PC13 <- loadings[order(loadings$Magnitude_PC13, decreasing = TRUE), ][1:10, ]

var_exp <- pca_ranks$sdev^2 /
  sum(pca_ranks$sdev^2) * 100

# Function to make a PCA plot
make_pca_plot <- function(
    sites,
    taxa,
    x_axis,
    y_axis,
    x_percent,
    y_percent,
    arrow_scale = 2
) {
  
  taxa_plot <- taxa
  
  taxa_plot$x_plot <- taxa_plot[[x_axis]] * arrow_scale
  taxa_plot$y_plot <- taxa_plot[[y_axis]] * arrow_scale
  
  ggplot() +
    
    # Sites
    geom_point(
      data = sites,
      aes(
        x = .data[[x_axis]],
        y = .data[[y_axis]]
      ),
      size = 4
    ) +
    
    geom_text(
      data = sites,
      aes(
        x = .data[[x_axis]],
        y = .data[[y_axis]],
        label = Site
      ),
      vjust = -1,
      size = 4
    ) +
    
    # Taxon arrows
    geom_segment(
      data = taxa_plot,
      aes(
        x = 0,
        y = 0,
        xend = x_plot,
        yend = y_plot
      ),
      arrow = arrow(
        length = unit(0.18, "cm")
      )
    ) +
    
    # Taxon labels
    geom_text(
      data = taxa_plot,
      aes(
        x = x_plot,
        y = y_plot,
        label = Taxon
      ),
      size = 3.2,
      vjust = -0.5
    ) +
    
    # Origin
    geom_hline(
      yintercept = 0,
      linetype = "dashed"
    ) +
    
    geom_vline(
      xintercept = 0,
      linetype = "dashed"
    ) +
    
    labs(
      x = paste0(
        x_axis,
        " (",
        round(x_percent, 1),
        "%)"
      ),
      y = paste0(
        y_axis,
        " (",
        round(y_percent, 1),
        "%)"
      )
    ) +
    
    theme_classic() +
    
    theme(
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10)
    )
}


# Create the two plots
p12 <- make_pca_plot(sites = sites, taxa = top10_PC12, x_axis = "PC1",
y_axis = "PC2", x_percent = var_exp[1], y_percent = var_exp[2])

p13 <- make_pca_plot(sites = sites, taxa = top10_PC13, x_axis = "PC1", y_axis = "PC3", x_percent = var_exp[1], y_percent = var_exp[3])

p12 + p13

ggsave("figures/Fig3_PCA.png",width = 20, height = 10, units = "cm")


# Loadings: which taxa contribute the most to the PC axes
loadings <- as.data.frame(pca_ranks$rotation)
contributions <- sweep(loadings^2, 2, colSums(loadings^2), "/") * 100
contributions$Taxon <- rownames(contributions)

contributions <- contributions[, c("Taxon", paste0("PC", 1:(ncol(contributions)-1)))]

contributions

top_contributions <- lapply(contributions[, -1], function(x) {contributions[order(x, decreasing = TRUE), c("Taxon", names(x))]})

top_contributions$PC1 #Clariidae, Trionychidae, Mochokidae, Polypterus, Synodontis
top_contributions$PC2 #Euthecodon, Bagridae, Libycosaurus, Polypterus, Claroteidae, Tetraodontidae
top_contributions$PC3 #Tetraodontidae, Synodontis, Claroteidae, Polypterus, Cichlidae
```

# Calculation of ecological scores

Based on the taxon ecological matrix, an ecological scores for each ecological category and assemblage was calculated. It first represents the relative contribution of taxa associated with a given ecological category to the assemblage. But, as ecological categories are represented by unequal numbers of taxa (e.g. 14 of the 22 taxa associated with rivers/channels, 12 with lakes, only 6 with swamps/marshes). As a consequence, ecological scores consistently show the strongest signals for rivers and lakes across all assemblages. To account for this bias, we applied a weighting factor that gives proportionally greater weight to ecological categories represented by fewer taxa. This prevents their ecological signal from being masked by categories represented by a larger number of taxa. Because the weighting factor differs among ecological categories, the resulting scores are not strictly standardised. They are therefore primarily intended to assess the relative expression of ecological signals among assemblages, while differences in scores among ecological categories should be interpreted as relative rather than strictly quantitative.

```{r}

calculate_eco_scores <- function(ranks, eco_matrix) {
  
  scores <- matrix(
    NA,
    nrow = ncol(ranks),
    ncol = ncol(eco_matrix)
  )
  
  rownames(scores) <- colnames(ranks)
  colnames(scores) <- colnames(eco_matrix)
  
  for (site in colnames(ranks)) {
    
    for (trait in colnames(eco_matrix)) {
      
      # Taxa associated with the ecological category
      taxa_trait <- eco_matrix[, trait] == 1
      taxa_trait[is.na(taxa_trait)] <- FALSE
      
      # Sum of weighted abundances
      numerator <- sum(
        ranks[taxa_trait, site],
        na.rm = TRUE
      )
      
      # Sum of weighted abundance of all taxa
      denominator <- sum(
        ranks[, site],
        na.rm = TRUE
      )
      
      scores[site, trait] <- numerator / denominator
    }
  }
  
  return(as.data.frame(scores))
}

eco_scores <- calculate_eco_scores(std_ranks_df, eco_matrix)

## Proportion of taxa associated to each ecological category
n_taxa_trait <- colSums(eco_matrix == 1, na.rm = TRUE)
eco_proportion <- n_taxa_trait / nrow(eco_matrix)

# Apply the correction
eco_scores_corrected <- sweep(eco_scores, 2, eco_proportion, "/")
eco_scores_corrected


# Visualisation with heatmap
# For visualisation, colour intensity was scaled independently within each assemblage using min–max normalization, so that the darkest and lightest colours represent the highest and lowest scores, respectively, within that assemblage. Colour intensity therefore highlights the relative ranking of ecological categories within each assemblage and should not be used for quantitative comparisons between assemblages.

# min–max normalization for each assemblage
heatmap_colors <- t(apply(as.matrix(eco_scores_corrected), 1, function(x) {(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))}))

my_palette <- colorRampPalette(c("#E8EEF3", "#577590", "#374A66"))(100)

pheatmap(
  heatmap_colors,
  color = my_palette,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = round(as.matrix(eco_scores_corrected), 2),
  number_format = "%.2f",
  legend = FALSE,
  border_color = "white",
  filename = "figures/Fig7_ecological_scores.png",
    width = 6,
  height = 4
)
```

# Data visualization

This section presents a reproducible workflow for visualizing the relative abundance of aquatic and terrestrial vertebrates at the studied Toros-Menalla fossil sites.

## Pie charts: Relative abundance of (semi-)aquatic vertebrates and terrestrial mammals per site

```{r}
# Prepare data for pie charts
TM_AqTer_long <- TM_AqTer %>%
  rownames_to_column("Taxon_Name") %>%
  pivot_longer(cols = -Taxon_Name, names_to = "Site", values_to = "Proportion")

colors_friendly <- c(
  "Actinopterygia" = "#2b3e5a", 
  "Crocodilia" = "#3a526e", 
  "Chelonia" = "#56748f", 
  "Hippopotamidae" = "#909db1", 
  "Anthracotheriidae" = "#c5c9d5",
  "Bovidae" = "#a6786e", 
  "Equidae" = "#ca7970", 
  "Giraffidae" = "#ee7770", 
  "Proboscidea" = "#f0967b", 
  "Suidae" = "#f2ba85", 
  "Other_terrestrial" = "#f2db8e"
)
taxon_order <- c("Actinopterygia", "Crocodilia", "Chelonia", "Hippopotamidae", "Anthracotheriidae",
                 "Bovidae", "Equidae", "Giraffidae", "Proboscidea", "Suidae", "Other_terrestrial")
TM_AqTer_long$Taxon_Name <- factor(TM_AqTer_long$Taxon_Name, levels = taxon_order)

# Plot pie charts
ggplot(TM_AqTer_long, aes(x = "", y = Proportion, fill = Taxon_Name)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  facet_wrap(~ Site) +
  scale_fill_manual(values = colors_friendly) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "Relative abundance of aquatic and terrestrial taxa per site")

ggsave("figures/Fig4_pie_charts.png")

```

## Barplots: Number of specimens per taxon and site

```{r}
# Long format
NISP_long <- NISP %>%
  rownames_to_column("Taxon") %>%
  pivot_longer(cols = -Taxon, names_to = "Site", values_to = "Count")

colors_aquatic <- c(
  "Alestinae" = "#567491", "Bagridae" = "#2d506d", "Cichlidae" = "#8fa4b5", 
  "Clariidae" = "#2d506d", "Claroteidae" = "#2d506d", "Crocodylus" = "#ada379",
  "Euthecodon" = "#ada379", "Gavialoidea" = "#ada379", "Gymnarchus" = "#567491",
  "Heterotis" = "#567491", "Hexaprotodon" = "#a6786e", "Hydrocynus" = "#567491",
  "Cyprinidae" = "#567491", "Lates" = "#8fa4b5", "Libycosaurus" = "#a6786e",
  "Mochokidae" = "#2d506d", "Mormyridae" = "#567491", "Polypterus" = "#567491",
  "Semlikiichthys" = "#8fa4b5", "Synodontis" = "#2d506d", "Tetraodontidae" = "#567491",
  "Trionychidae" = "#ada379")

taxon_order <- c(
  "Alestinae", "Hydrocynus", "Heterotis", "Gymnarchus", "Mormyridae", "Polypterus", 
  "Cyprinidae", "Tetraodontidae", "Bagridae", "Clariidae", "Claroteidae", 
  "Mochokidae", "Synodontis", "Cichlidae", "Lates", "Semlikiichthys", 
  "Crocodylus", "Euthecodon", "Gavialoidea", "Trionychidae", 
  "Hexaprotodon", "Libycosaurus")

NISP_long$Taxon <- factor(NISP_long$Taxon, levels = taxon_order)

ggplot(NISP_long, aes(x = Taxon, y = Count, fill = Taxon)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors_aquatic) +
  facet_wrap(~ Site, ncol = 1) +
  theme_minimal() +
  labs(title = "Number of specimens per taxon and site",
       x = "Taxa", y = "Number of specimens") +
  ylim(0, 100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

ggsave("figures/Fig6_barplots_NISP.png",width = 10, height = 24, units = "cm")

```

## Barplots: Number of specimens per taxon and site

```{r}
std_ranks_df_long <- std_ranks_df %>%
  rownames_to_column("Taxon") %>%
  pivot_longer(cols = -Taxon, names_to = "Site", values_to = "Count")

std_ranks_df_long$Taxon <- factor(std_ranks_df_long$Taxon, levels = taxon_order)

ggplot(std_ranks_df_long, aes(x = Taxon, y = Count, fill = Taxon)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors_aquatic) +
  facet_wrap(~ Site, ncol = 1) +
  theme_minimal() +
  labs(title = "Standardised taxonomic ranks of abundance",
       x = "Taxa", y = "Standardised abundance") +
  ylim(0, 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

ggsave("figures/Fig6_barplots_ranks.png",width = 10, height = 24, units = "cm")
```
